---
title: "NYPD Shooting Incident Data Report"
author: "Ryan Masad"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
```

# NYPD Shooting Incident Data Report

The NYPD Shooting Incident Data provides insight into cases of shooting since 2006. The data set provides information on when the event took place, where it took place, whether it resulted in a death, and any information that is available about the perpetrator and the victim. The question that I want to dive into is about how shootings and murders correlate within various boroughs. 

### Loading the Data

Start by loading the data from the New York shooting report and examining the data for any immediate fixes.

```{r get_shooting_data}
url_file = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
ny_shooting = read.csv(url_file)
ny_shooting %>% head(5)
summary(ny_shooting)
```

### Cleaning the Data

After initial examination of the above data the following initial cleaning were identified.

1.  Convert Date to a Date type variable
2.  Create a column with Year, Month, and Monthly Date
    a.  This is needed because there isn't a shooting every day. So tracking trends over time seems more useful.
3.  Convert STATISTICAL_MURDER_FLAG into a binary variable so we can sum it up to see how many murders occurred
4.  Clean up the data set and remove unneeded columns. INCIDENT_KEY, TIME, PRECINCT, and JURISDICTION CODE do not seem necessary

```{r ny_shooting_data_cleaning}
ny_shooting = ny_shooting %>% 
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y")) %>% 
  mutate(YEAR = year(OCCUR_DATE)) %>%  
  mutate(MONTH = month(OCCUR_DATE)) %>%  
  mutate(MONTHLY_DATE = floor_date(OCCUR_DATE, unit="month")) %>%
  mutate(STATISTICAL_MURDER_FLAG = ifelse(STATISTICAL_MURDER_FLAG == "true",1,0)) %>%
  select(OCCUR_DATE, MONTHLY_DATE, YEAR, MONTH, BORO, LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC:VIC_RACE)

ny_shooting %>% head(5)
```

### Population Data Import

After Cleaning the data, Population seems like it could aid in analyzing the following data. So I found a population file off the same site. I chose to average population for 2010 and 2020 as there weren't records for every year and I think the average will give a decent idea of the population. Using just 2020 or 2010 would be valid as well.

```{r get_boro_population_data}
boro_file = "https://data.cityofnewyork.us/resource/xywu-7bv9.csv"
boro_data = read.csv(boro_file)

boro_data = boro_data %>% 
  mutate(borough = trimws(str_to_upper(borough))) %>%
  mutate(population = round((X_2020+X_2010)/2,0)) %>%
  select(borough,population)
boro_data
```

### Join the Data

Here I am joining in the Population data to the NY_shooting database.

```{r join_shooting_to_population_data}
ny_shooting = left_join(ny_shooting, boro_data, by = join_by("BORO" == "borough"))
ny_shooting %>% head(5)
```

### New York Shootings by Month

Starting off with a simple analysis, I wanted to see how shootings and murders trended for the entirety of New York. I am looking at the data by month.The first thing that stands out to me, is the seasonal nature to shootings. The data spikes in some seasons and then drops. It also looks like there was a trend toward less shooting up until 2020. Perhaps Covid had a severe effect on crime in New York.

```{r ny_shootings_by_month}
shootings_by_month = ny_shooting %>% 
  group_by(MONTHLY_DATE,MONTH) %>%
  summarise(shootings = length(OCCUR_DATE), murders = sum(STATISTICAL_MURDER_FLAG)) 
shootings_by_month
```

```{r plot_shootings_by_month, echo=FALSE}
shootings_by_month %>%
  ggplot(aes(x=MONTHLY_DATE, y=shootings)) +
  geom_line(aes(color = "shootings")) + 
  geom_point(aes(color="shootings")) +
  geom_line(aes(y=murders, color="murders")) + 
  geom_point(aes(y=murders, color="murders")) +
  ggtitle("Shootings Trend") +
  xlab("Month") +
  ylab("Shootings") +
  theme_classic()
```

### Diving Deeper into Shootings By Month

Now i want to look at the average \# of shootings over the last 15 years by month. When diving into this, you can see a peak during the summer and it dropping during the winter. This makes sense as people tend to stay home when it snows likely resulting in lower crime rates.

```{r ny_shootings_by_month_no_year}
shootings_by_month_no_year = shootings_by_month %>% 
  group_by(MONTH) %>%
  summarise(shootings = mean(shootings), murders = mean(murders)) 
shootings_by_month_no_year
```

```{r plot_shootings_by_month_no_year, echo=FALSE}
shootings_by_month_no_year %>% 
  ggplot(aes(x = MONTH, y = shootings)) +
  geom_bar(stat = "identity", fill = "blue") +
  ggtitle("Shootings by Month",) +
  xlab("Month") +
  ylab("Shootings") +
theme_classic()
```

### Shootings by Boro

Next I want to look at Shootings by Year by BORO to focus more on the decline in shootings over the years followed by the spike. I wanted to see how boroughs see crime rates differently. I start by graphing shootings and shootings per million by boroughs. This shows that while Brooklyn has the most shootings, Bronx has the highest shootings per million people. The other 3 boroughs all have similar shootings per million. Murders show similar results.

```{r shootings_by_boro_by_year}
shootings_by_boro = ny_shooting %>% 
  group_by(BORO, YEAR) %>%
  summarise(shootings = length(OCCUR_DATE), murders = sum(STATISTICAL_MURDER_FLAG), population = max(population)) %>%
  mutate(shootings_per_pop = shootings/population*1000000) %>%
  mutate(murders_per_pop = murders/population*1000000)
shootings_by_boro %>% head(5)
```

```{r plot_shootings_by_boro_by_year, echo=FALSE}
shootings_by_boro %>% 
  ggplot(aes(x = YEAR, y = shootings, group = BORO, color = BORO)) +
  geom_line() +
  geom_point() +
  ggtitle("Shootings by Borough and Year") +
  xlab("Year") +
  ylab("Number of Shootings") +
  theme_classic()

shootings_by_boro %>% 
  ggplot(aes(x = YEAR, y = shootings_per_pop, group = BORO, color = BORO)) +
  geom_line() +
  geom_point() +
  ggtitle("Shootings per Million by Borough and Year") +
  xlab("Year") +
  ylab("Shootings per Million") +
  theme_classic()

shootings_by_boro %>% 
  ggplot(aes(x = YEAR, y = murders, group = BORO, color = BORO)) +
  geom_line() +
  geom_point() +
  ggtitle("Murders by Borough and Year") +
  xlab("Year") +
  ylab("Number of Murders") +
  theme_classic()

shootings_by_boro %>% 
  ggplot(aes(x = YEAR, y = murders_per_pop, group = BORO, color = BORO)) +
  geom_line() +
  geom_point() +
  ggtitle("Murders per Million by Boro and Year") +
  xlab("Year") +
  ylab("Murders per Million") +
  theme_classic()
```


### Linear Model of Shootings against Murders

Now i want to verify that murders per million people is a function of Shootings per million people. We also found out that there is high correlation between shootings and murders with R^2^ being ~ 97%.

```{r model_shootings_by_boro_by_year}
model = lm(shootings_per_pop ~ murders_per_pop + factor(BORO) + factor(YEAR), data = shootings_by_boro)
summary(model)

predictions = predict(model, newdata = shootings_by_boro)

shootings_by_boro$predicted_shootings_per_pop = predictions
shootings_by_boro %>% head(5)
```

Finally let's plot this to show a strong correlation between the predictions and the actual values.

```{r plot_model_shootings_by_boro_by_year, echo=FALSE}
ggplot(shootings_by_boro, aes(x = predicted_shootings_per_pop, y = shootings_per_pop)) +
  geom_point(aes(color = BORO), size = 3) +  # Color points by BORO
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Add a linear regression line
  ggtitle("Observed vs Predicted Shootings per Population") +
  xlab("Predicted Shootings per Population") +
  ylab("Observed Shootings per Population") +
  theme_classic()
```


### Continuation. 
Now that we know murders and shootings share similar rates across boroughs, I want to dive deeper into the victims, by looking at Race and Sex involved in shootings by borough. 

```{r ny_shootings_by_month_by_victim}
shootings_by_victim = ny_shooting %>% 
  group_by(BORO,YEAR,MONTHLY_DATE,MONTH,VIC_SEX, VIC_RACE) %>%
  summarise(shootings = length(OCCUR_DATE), murders = sum(STATISTICAL_MURDER_FLAG)) 
shootings_by_victim 
```
```{r plot_shootings_by_sex_boro, echo=FALSE}
shootings_by_victim %>% 
  ggplot(aes(x = BORO, y = shootings, fill=VIC_SEX)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Shootings by Sex and Borough",) +
  xlab("Borough") +
  ylab("Shootings") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust=1))
```
```{r table_shootings_by_sex_boro, echo=FALSE}
shootings_by_victim %>% 
  group_by(BORO, VIC_SEX) %>%
  summarise(shootings = sum(shootings)) %>%
  spread(key = VIC_SEX, value = shootings) %>%
  mutate(total = M + F, 
         percent_male = (M / total) * 100)
```

```{r plot_shootings_by_race_boro, echo=FALSE}
shootings_by_victim %>% 
  ggplot(aes(x = BORO, y = shootings, fill=VIC_RACE)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Shootings by Race and Borough",) +
  xlab("Borough") +
  ylab("Shootings") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust=1))
```
```{r plot_shootings_by_sex_race, echo=FALSE}
shootings_by_victim %>% 
  ggplot(aes(x = VIC_RACE, y = shootings, fill=VIC_SEX)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Shootings by Race and Sex") +
  xlab("Race") +
  ylab("Shootings") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust=1))
```
```{r table_shootings_by_sex_race, echo=FALSE}
shootings_by_victim %>% 
  group_by(VIC_SEX, VIC_RACE) %>%
  summarise(shootings = sum(shootings)) %>%
  spread(key = VIC_SEX, value = shootings) %>%
  mutate(total = M + F, 
         percent_male = (M / total) * 100)
```

### Conclusion

We have seen heavy correlation between shooting and murders and strong seasonality with the data.While Male victims occur much more frequently, Female victims nearly match male victims in Staten Island, which posted the lowest number of shootings.Additionally, while Race also showed much higher shootings involving Black Victims, White Hispanics were also incredibly high in Bronx and Staten Island. 

Some areas of bias, that affect this data set are:

1.  The spike of violent crimes write around 2020. This may not be representative of a Normal New York City, and may lead to a bad fitting model. We could remove this data, but that would also likely remove the return to pre improvement that New York was achieving.
2.  Races and Age also are an area of Bias. with race and age being included, you could attempt to manipulate the data into making certain races or age groups look worse.
    a.  I chose to not analyze race and age for my analysis, but this could lead to false conclusions.
3.  Boroughs to shooting ratio. Some boroughs have much higher shootings. This could make a boroughs look significantly worse or better even if it isn't. I chose to normalize the data by pulling in population to find the average shootings per million people.
